---
title: "Bayesian Priors Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Priors Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = 'figures/',
  out.width="100%",
  fig.align = "center",
  tidy=TRUE,
  tidy.opts=list(width.cutoff=40)
)
```

## Overview

This is the single maintained end-to-end worked example for PriorRhythm. It
covers the two supported ways of using the package:

1. **Primary workflow — using the Shiny app.** The Shiny app (`run_app()`)
   is a **priors-only runtime**: by default it loads the reviewed prior
   artifact bundled with the package and does **not** build priors from raw
   historical data during an app session. Users launch the app, upload
   target-study data, select a compatible prior, and review results. This is
   the supported workflow for the poster-companion release.
2. **Advanced workflow — building custom priors.** Advanced users who need a
   prior that is not already in the bundled artifact can build one offline
   from historical data with `chain_priors_from_data()` and supply the
   resulting bundle to the app via `run_app(prior_file = ...)`. This section
   is intentionally concise and API-focused; it does not explain JAGS/MCMC
   internals.

All data used in this vignette is **synthetic demonstration data only**,
shipped with the package under `inst/extdata/`. It does not represent
historical production study data.

## Load package

```{r setup}
library(PriorRhythm)
```

```{r load-data, include=FALSE}

# Load synthetic historical data from inst/extdata/
SynHist <- system.file("extdata", "SyntheticHistorical",
  package = "PriorRhythm",
  mustWork = TRUE
)

Historical_data_long <- readRDS(file = file.path(SynHist, "Syn_long.RDS")) |>
  dplyr::rename(name = "parameter")

# Load synthetic test data from inst/extdata/
SynTest <- system.file("extdata", "SyntheticTest",
  package = "PriorRhythm",
  mustWork = TRUE
)

Raw_Test_Study <- readRDS(file = file.path(SynTest, "Syn_test.RDS"))

```

# Primary workflow: the Shiny app

## Launching the app

```{r eval=FALSE}
run_app()
```

By default `run_app()` resolves the bundled prior artifact shipped at
`inst/extdata/priors/list_historical_prior.RDS` — no raw historical data and
no MCMC sampling occur during the app session. Advanced users can instead
supply a custom, pre-built, validated bundle with `run_app(prior_file =
"path/to/bundle.RDS")`.

## Uploading target-study data

The **File upload** tab accepts CSV and TSV files (see `accept` in
`mod_file_upload_sidebar_ui()`). Uploaded data is read with `load_study_data()`,
which requires a **wide**-format target-study data frame.

The app shows reviewed File Upload guidance in an expandable **File
requirements and workflow** panel above the uploaded-data preview, and shows
reviewed species/endpoint display labels where package metadata is available,
while preserving raw prior-bundle keys internally for selection and lookup.

### Required columns

-   animal_col = individual animal code
-   treatment_col = treatment level codes
-   period_col = period code
-   time_col = time bins
-   y_col = assay data

```{r}
Raw_Test_Study |>
  dplyr::select(dplyr::any_of(c(
    "animal_code",
    "treatment_code",
    "period_code",
    "time", "time1", "time2",
    "value")))

```

`time_bins()` reformats the time column into `time1`/`time2`; the app applies
this automatically once the time column and delimiter are supplied in the
sidebar.

```{r}
Test_Study <- Raw_Test_Study |>
  time_bins(time_col = "time", delimiter = "to")

Test_Study |>
  dplyr::select(dplyr::any_of(c(
    "animal_code",
    "treatment_code", "period_code",
    "time", "time1", "time2", "value")))

```

## Selecting a compatible prior

The app lets users browse the bundled priors by species and endpoint, but it
does **not** automatically verify that a selected prior is appropriate for
the uploaded data. Users are responsible for selecting a prior whose
**species**, **endpoint definition**, and **measurement unit** all match the
target-study data before running an analysis; selecting a mismatched prior
will produce a result without a warning.

## Testing target-study data against a prior

`bayesian_test_example()` calculates posteriors using the selected prior and
a weak default prior, so results can be compared side by side.

```{r eval=FALSE}
# Bundled prior for the first available species/endpoint, as would be
# selected in the app
Bundle <- readRDS(
  system.file("extdata", "priors", "list_historical_prior.RDS",
    package = "PriorRhythm", mustWork = TRUE))

# Bundles use either a versioned envelope (`Bundle$priors`) or a legacy flat
# shape (species keys at the top level); resolve either form here.
priors_tree <- if (!is.null(Bundle$priors)) Bundle$priors else Bundle

# Species/endpoint keys available in the bundled artifact (e.g. "dog"/"nhp"
# species with endpoints such as "q_tc", "hr", "dbp")
species  <- names(priors_tree)[1]
endpoint <- names(priors_tree[[species]])[1]
selected_prior <- priors_tree[[species]][[endpoint]]$pa

# running analysis
Test_out1 <- bayesian_test_example(
  input_tdat    = Test_Study,
  animal_col    = "animal_code",
  treatment_col = "treatment_code",
  period_col    = "period_code",
  time1_col     = "time1",
  time2_col     = "time2",
  y_col         = "value",
  n_exclude     = 1,
  pa            = selected_prior
)
```

### Credible intervals plot

Both branches are fitted with `MCMCglmm`; the comparison is which prior is
used, not whether the target data are complete. `Historical prior — full
data` (`Informative`) uses the selected historical prior on the target
dataset. `Weak-prior comparator — full data` (`Non_informative`) fits the
same model with `MCMCglmm`'s weak default prior (no `prior` argument
supplied) instead of the historical prior.

```{r CI_Plot, eval=FALSE, fig.height=10, fig.width=12}
CIPlots <- ci_plots(Test_out1,
  animal_col    = "animal_code",
  treatment_col = "treatment_code",
  period_col    = "period_code",
  time1_col     = "time1",
  time2_col     = "time2",
  y_col         = "y",
  threshold     = 10,
  time_code_col = "time_code",
  size          = 1,
  input_method  = c(
    "Weak-prior comparator \u2014 full data" = "Non_informative",
    "Historical prior \u2014 full data"      = "Informative"
  ))

CIPlots +
  ggplot2::theme_classic(base_size = 16) +
  ggplot2::theme(
    plot.margin     = grid::unit(c(1,1,1,1), "cm"),
    legend.position = "bottom",
    legend.box      = "vertical",
    legend.margin   = ggplot2::margin(),
    aspect.ratio    = 1
  )

```

Credible intervals summarize the posterior distribution of the treatment
effect under each model; they are descriptive summaries of borrowed-strength
estimation for this dataset, not a determination of clinical or regulatory
significance.

### Probability statements

```{r Prob_statement, eval=FALSE, fig.height=10, fig.width=10}
prob_statement_plots(Test_out1,
  threshold    = 10,
  input_method = c(
    "Weak-prior comparator \u2014 full data" = "Non_informative",
    "Historical prior \u2014 full data"      = "Informative"
  )) +
  ggplot2::theme_classic(base_size = 16) +
  ggplot2::theme(
    plot.margin     = grid::unit(c(1,1,1,1), "cm"),
    legend.position = "bottom",
    legend.box      = "vertical",
    legend.margin   = ggplot2::margin(),
    aspect.ratio    = 1
  )

```

`prob_statement_plots()` reports the posterior probability that the
treatment effect exceeds a user-specified threshold. This is a probability
statement about the fitted model given the supplied prior and target data —
it is not, by itself, a regulatory conclusion or a substitute for expert
review.

## Prior-bundle diagnostics

Packaged prior bundles record non-confidential, chain-aware convergence
evidence (R-hat, effective sample size, and pass/acceptable/failed status)
captured during the controlled Stage 1 build. The app's runtime analyses use
the pooled posterior draws (`pa`) from the bundle; chain identity is not
retained in `pa`, so the app **does not**, and cannot, recompute Stage 1
chain-aware diagnostics at runtime. Reviewing convergence is part of the
controlled build/refresh process for a prior bundle, not something the app
repeats for every analysis.

# Advanced workflow: building custom priors from historical data

This section is intentionally brief. It is only needed if the bundled prior
artifact does not already contain a compatible prior for the
species/endpoint you need.

## Historical data requirements and format

Data must be in **long** format with the following columns:

-   study_col = `study_code` column
-   animal_col = individual animal code
-   treatment_col = treatment level codes
-   period_col = period code
-   time_col = time bins
-   assay_names_col = name column generated by pivot_longer
-   assay_values_col = value column generated by pivot longer

```{r Historical data}

Historical_data_long |>
  dplyr::select(
    dplyr::any_of(c(
      "study_code", "animal_code",
      "time",
      "treatment_code",
      "period_code",
      "name", "value")))

```

The time bin column must be split into `time1`/`time2` before calling
`chain_priors_from_data()`:

```{r}
Updated_Historical_data_long <- Historical_data_long |>
  time_bins(time_col = "time", delimiter = "to")

Updated_Historical_data_long |>
  dplyr::select(dplyr::any_of(c(
    "study_code", "animal_code",
    "time", "time1", "time2",
    "treatment_code",
    "period_code",
    "name", "value")))

```

## Building priors

`chain_priors_from_data()` requires `alpha0_pop_mean` — either supplied
explicitly (as below) or via a `study_config()` endpoint configuration. There
is no default; a `NULL` value aborts with an informative error.

```{r eval=FALSE}
# may take a few minutes — runs JAGS MCMC
List_of_Priors <- Updated_Historical_data_long |>
  chain_priors_from_data(
    treatment_col    = "treatment_code",
    study_col        = "study_code",
    animal_col       = "animal_code",
    period_col       = "period_code",
    assay_names_col  = "name",
    assay_values_col = "value",
    assay_name       = c("assay1", "assay2"),
    alpha0_pop_mean  = 250)

```

Equivalently, using `study_config()` to supply `alpha0_pop_mean` per
endpoint:

```{r eval=FALSE}
cfg <- study_config(
  endpoints = list(
    assay1 = list(alpha0_pop_mean = 250),
    assay2 = list(alpha0_pop_mean = 70)
  ))

List_of_Priors <- Updated_Historical_data_long |>
  chain_priors_from_data(
    treatment_col    = "treatment_code",
    study_col        = "study_code",
    animal_col       = "animal_code",
    period_col       = "period_code",
    assay_names_col  = "name",
    assay_values_col = "value",
    assay_name       = c("assay1", "assay2"),
    config           = cfg)

```

The output of this function is a named list matching the values supplied to
the `assay_name` argument.

```{r eval=FALSE}
names(List_of_Priors)

```

### Plot assay1 prior

```{r Plot_prior_assay1, eval=FALSE, fig.height=8, fig.width=12}
chain_prior_plot(
  interval_start_times = c(0,3,6,12,18),
  interval_end_times   = c(3,6,12,18,24),
  doses                = seq(0,3),
  prior_list           = List_of_Priors,
  prior_name           = "assay1") |>
  ggpubr::ggarrange(plotlist = .)

```

### Plot assay2 prior

```{r Plot_prior_assay2, eval=FALSE, fig.height=8, fig.width=12}
chain_prior_plot(
  interval_start_times = c(0,3,6,12,18),
  interval_end_times   = c(3,6,12,18,24),
  doses                = seq(0,3),
  prior_list           = List_of_Priors,
  prior_name           = "assay2") |>
  ggpubr::ggarrange(plotlist = .)

```

A bundle intended for use with the app must be assembled with
`make_runtime_prior_bundle()` (which records the required convergence
`diagnostics` for each endpoint) before it is passed to `run_app(prior_file =
...)`; see that function's help page for the required schema.

# Limitations and reproducibility

- All example data in this vignette is **synthetic**; it is generated for
  demonstration only and is not derived from historical production studies.
- The Shiny app is a priors-only runtime: it consumes an already-built,
  validated prior bundle and never fits a Stage 1 JAGS model during an app
  session.
- The current target-study upload contract accepts CSV and TSV files only.
- The app uses reviewed species/endpoint display labels and File Upload
  guidance from packaged YAML when available, but valid bundled or trusted
  custom priors remain usable even when those display metadata entries are
  absent.
- Users are responsible for confirming that a selected prior's species,
  endpoint definition, and measurement unit are compatible with the
  uploaded target-study data; the app does not enforce this automatically.
- Credible intervals and probability statements reported by this package
  are descriptive statistical summaries; they are not, by themselves,
  regulatory determinations.
- Convergence diagnostics for a prior bundle are recorded once, during the
  controlled build of that bundle, and are not recomputed by the app at
  analysis time.
